#This is a personal activity data analysis markdown file.

##1. Loading and preprocessing the data

```{r}
getwd()
setwd("/Users/hui/datasciencecourses/reproducibleResearch/RepData_PeerAssessment1")
getwd()

activity <- read.csv("activity.csv",header=TRUE)
str(activity)

cleanActivity <- activity[!is.na(activity$steps), ]
```

##2.What is mean total number of steps taken per day?
### Make histogram to show  total steps each day
```{r}
library(plyr)
# group values by date
groupedByDate <- ddply(cleanActivity, ~date, summarise, sum = sum(steps))
#groupedByDate
#Make a histogram of total number of steps taken each day
hist(groupedByDate$sum, xlab = "Total Steps", main = "Histogram total steps each day", 
     col = "black")
```

### What is the mean total number of steps?
```{r}

# Calculate mean value
mean(groupedByDate$sum)
```

### What is the median total number of steps?
```{r}

# Calculate median value
median(groupedByDate$sum)
```

##3. What is the average daily activity pattern?
```{r}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average #number of steps taken, averaged across all days (y-axis)
groupedByInterval <- ddply(cleanActivity, ~interval, summarise, mean = mean(steps))
with(groupedByInterval, plot(interval, mean, type = "l", ylab = "Average number of steps", 
                             xlab = "Interval", main = "Average daily activity", col = "black"))
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# first find max value
maxVal <- max(groupedByInterval$mean)
print(maxVal)

# locate the line containing this value
maxLine <- groupedByInterval[groupedByInterval$mean == maxVal, ]
# find the interval
maxInterval <- maxLine$interval
print(maxInterval)

```

##4. Imputing missing values
### How many missing values are there in the dataset?
```{r}
#  Calculate and report the total number of missing values in the dataset (i.e. the total number #of rows with NAs)
sum(is.na(activity$steps))
```

### Devise a strategy for filling in all of the missing values in the dataset and then plot. I use the mean for 5-minute interval.
```{r}
# create a new dataset
newdata <- activity
# locate tha NAs
missingsteps <- is.na(newdata$steps)
# convert interval(s) to factor(s)
newdata$interval <- factor(newdata$interval)
groupedByInterval$interval <- factor(groupedByInterval$interval)

# fill newdata on missing values wherever we have a missing value, fill it
# from groupedByInterval$mean column (steps) where newdata$interval ==
# groupedByInterval$interval (intervals are factors)
#newdata[missingsteps, "interval"]
class(newdata[missingsteps, "interval"])
newdata[missingsteps, "steps"] <- groupedByInterval[newdata[missingsteps, "interval"], 
                                                    "mean"]

# group values by date
groupedByDate2 <- ddply(newdata, ~date, summarise, sum = sum(steps))

# construct the histogram
hist(groupedByDate2$sum, xlab = "Total Steps", main = "Histogram total steps per day", 
     col = "darkmagenta")
```

###After filling missing values, what is the mean total number of steps?
```{r}
# Calculate mean value
mean(groupedByDate2$sum)
```

###After filling missing values, what is the median total number of steps?
```{r}
# Calculate median value
median(groupedByDate2$sum)

```

##5. Are there differences in activity patterns between weekdays and weekends?
```{r}

# add a new column containing day of week
# change factor type to Date then use function weekdays
newdata$weekday = weekdays(as.Date(newdata$date, format="%Y-%m-%d"))
# add a new column containing either Weekday OR Weekend
newdata$weekday.type <- ifelse(newdata$weekday == "Saturday" | newdata$weekday == 
                                   "Sunday", "Weekend", "Weekday")
# convert column to factor
newdata$weekday.type <- factor(newdata$weekday.type)

# make a new dataset grouping data by interval and weekday.type
groupedBy.Interval.WeekDay <- ddply(newdata, ~interval + weekday.type, summarise, 
                                    mean = mean(steps))

# convert to characters first
groupedBy.Interval.WeekDay$interval <- as.numeric(as.character(groupedBy.Interval.WeekDay$interval))
library(lattice)
xyplot(mean ~ interval | weekday.type, groupedBy.Interval.WeekDay, type = "l", 
       layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")


```

